Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-04 paper/main.tex
Updates

This CI testing

@@

In this CI testing procedure, users test whether the implied CIs of the DAG
hold in the dataset or not using statistical CI tests. Based on the results of
these statistical tests, we can accordingly make modifications to the current
model to fix the violations. However, this CI based testing also has a few
drawbacks: 1) CIs are implied only by missing edges in the model, so we can not
test for already existing edges. 2) In high sample size scenarios, p-values can
suggest a significant association between variables even if they have a very
weak relationship. 3) Number of implied CIs can be large making it difficult to
manually go through each of them; for example, in the moderately sized alarm
network with $ 37 $ node, there are $287$ implied CIs.

@@

To tackle these issues with CI based testing, we take inspiration from
\emph{modification indices} in specification search. The modification indices
method tries to identify and rank changes in the model such that modifications
that improve the model fit the most are ranked higher.This ranking allows
researchers to focus first on the suggestions that can improve the model the
most and while integrating their expertise to choose the most optimal
modifications. Analogous to this, we develop a method for ranking CI test
violations. This ranking method utilizes both p-values and a measure of
association that we develop in this paper. Using this ranking we present a
procedure for iteratively modifying the DAG while integrating expert knowledge
into it.

\begin{figure}
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=1]{figures.pdf}
	\caption{$ N = 400 $}
\end{subfigure}%
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=2]{figures.pdf}
	\caption{$ N = 800$ }
\end{subfigure}
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=5]{figures.pdf}
	\caption{$ N = 400 $ }
\end{subfigure}%
\begin{subfigure}{0.25\textwidth}
	\includegraphics[page=6]{figures.pdf}
	\caption{ $ N = 800$ }
\end{subfigure}
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=3]{figures.pdf}
	\caption{ $ N = 400$ }
\end{subfigure}%
\begin{subfigure}{0.25\textwidth}
	\includegraphics[page=4]{figures.pdf}
	\caption{$ N = 800$ }
\end{subfigure}

@@

Our main contributions in this paper are as follows:
\begin{enumerate}
	\item We present a measure of conditional association for mixed data
		based on canonical correlations. This measure of association
		generalizes some of the commonly used special case measures of 
		association to mixed data.
	\item Using this conditional association measure, we present a procedure to
		manually construct DAGs by ranking violation of implied CIs in the 
		model. This iterative procedure allows users to integrate their
		domain knowledge in the model construction procedure.
	\item Lastly, to allow researchers to easily apply this method on their
		datasets, we present a web-browser based interactive tool as well
		as implementation in a Python package.
\end{enumerate}

The rest of the paper is structured as follows. In
Section~\ref{sec:background}, we give a background on the commonly used
measures of association for various data types. In
Section~\ref{sec:mixed_association}, we present our generalized measure of
conditional association for mixed data. Section~\ref{sec:modification} provides
details on the procedure for using the measure of association for constructing
DAGs. Section~\ref{sec:web} presents our web-browser based tool and lastly in
Section~\ref{sec:empirical}, we show some empirical results to compare this
manual approach to automated algorithms.

@@

\begin{figure}
	\begin{subfigure}{0.17 \textwidth}
		\includegraphics[page=1]{example.pdf}
		\caption{True DAG}
	\end{subfigure}%
	\begin{subfigure}{0.17 \textwidth}
		\includegraphics[page=2]{example.pdf}
		\caption{No edges}
	\end{subfigure}%
	\begin{subfigure}{0.17 \textwidth}
		\includegraphics[page=3]{example.pdf}
		\caption{Orient $ X_1 \rightarrow X_3 $}
	\end{subfigure}
	\begin{subfigure}{0.17 \textwidth}
		\includegraphics[page=4]{example.pdf}
		\caption{Orient $ X_2 \rightarrow X_3 $}
	\end{subfigure}%
	\begin{subfigure}{0.17 \textwidth}
		\includegraphics[page=5]{example.pdf}
		\caption{Orient $ X_3 \rightarrow X_4 $}
	\end{subfigure}%
	\begin{subfigure}{0.17 \textwidth}
		\includegraphics[page=6]{example.pdf}
		\caption{Orient $ X_3 \rightarrow X_5 $}
	\end{subfigure}
	\caption{An example showing a greedy approach for causal discovery using the mixed data measure of association. We start with an empty graph (b) where all the variables in the model are associated with each other. The strength of association is denoted using the thickness of the dotted edges. We select the pair of variables with the highest association, for example, $X_1$ and $X_3$, and orient it based on our domain knowledge. We continue the process till we have a DAG.
	}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[scale=0.4]{../code/plots/web_tool_full.png}
	\caption{A screenshot of the web tool for constructing the model. Users
		can upload their dataset after which the tool creates an empty
		graph and shows all pair of variables which are associated in
		the model using undirected red edges with the strength of
		association represented using edge width. Users can then add
		edges to the model (shown in green). Unnecessary edges are shown in 
		black.}
\end{figure}

@@

\begin{figure}
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/shd_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/sid_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\caption{Comparison of PC and Hill-Climb Search algorithms against
		manually drawn DAGs using the assistance method. As PC and
		Hill-Climb Search return the Markov Equivalence Class (MEC), we
		use the best and worst scoring orientation of the MEC to get
		the range of values. The human drawn values are done for
		multiple accuracy values of the oracle.}
\end{figure}

\begin{figure}
	\begin{subfigure}{0.25\textwidth}
		\centering
		\includegraphics{../code/fig4/unexplained_effect.pdf}
		\caption{}
	\end{subfigure}%
	\begin{subfigure}{0.25\textwidth}
		\centering
		\includegraphics{../code/fig4/ll.pdf}
		\caption{}
	\end{subfigure}
	\caption{Plots showing improvements in fit of the model over iterations
	on the Adult income dataset. For the expert simulation an LLM is used
	for the edge orientations. (a) Shows using total unexplained effect that is the
	sum of effect size between variables that don't have an edge between them. (b)
	Shows the log-likelihood of the overall model.
	}
\end{figure}

@@

\begin{figure}
	\centering
	\begin{subfigure}{0.5\textwidth}
		\begin{Verbatim}[fontsize=\footnotesize]
You are an expert in Social Science. Following
are the descriptions of two variables:

>>>>

However, this CI testing approach also has a few drawbacks: 1) CIs are implied
only by missing edges in the model, so we can not test whether the existing
edges are correct or not. 2) The decision whether a CI holds in the data or not
is based on a p-value threshold. But we can get significant p-values for even
very weak relationship if the sample size is high. 3) Lastly, the number of
implied CIs can be large making it difficult to go through all of them; for 
example, in the moderately sized alarm network \citep{Beinlich1989} with $ 37 $
nodes, there are $ 287 $ implied CIs.

@@

% In this CI testing procedure, users test whether the implied CIs of the DAG
% hold in the dataset or not using statistical CI tests. Based on the results of
% these statistical tests, we can accordingly make modifications to the current
% model to fix the violations. However, this CI based testing also has a few
% drawbacks: 1) CIs are implied only by missing edges in the model, so we can not
% test for already existing edges. 2) In high sample size scenarios, p-values can
% suggest a significant association between variables even if they have a very
% weak relationship. 3) Number of implied CIs can be large making it difficult to
% manually go through each of them; for example, in the moderately sized alarm
% network with $ 37 $ node, there are $287$ implied CIs.

@@

To tackle these issues with CI based testing, we take inspiration from
\emph{modification indices} in specification search. The modification indices
method tries to identify and rank changes in the model such that modifications
that improve the model fit the most are ranked higher. This ranking allows
researchers to focus first on the suggestions that can improve the model the
most and while integrating their expertise to choose the most optimal
modifications. Analogous to this, we develop a procedure for ranking CI test
violations allowing researchers to focus on the worst violations in the model.
This ranking method utilizes both p-values and a measure of conditional
association. Utilizing this procedure researchers can iteratively construct
DAGs while integrating their domain knowledge in the model building procedure.

\begin{figure}[t!]
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=1]{figures.pdf}
	\caption{$ N = 400 $}
\end{subfigure}%
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=2]{figures.pdf}
	\caption{$ N = 800$ }
\end{subfigure}
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=5]{figures.pdf}
	\caption{$ N = 400 $ }
\end{subfigure}%
\begin{subfigure}{0.25\textwidth}
	\includegraphics[page=6]{figures.pdf}
	\caption{ $ N = 800$ }
\end{subfigure}
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=3]{figures.pdf}
	\caption{ $ N = 400$ }
\end{subfigure}%
\begin{subfigure}{0.25\textwidth}
	\includegraphics[page=4]{figures.pdf}
	\caption{$ N = 800$ }
\end{subfigure}

@@

Our main contributions in this paper are as follows:
\begin{enumerate}
	\item We propose a novel measure of conditional association for mixed data
		based on canonical correlations. This measure of association
		generalizes some of the commonly used special case measures of 
		association to mixed data.
	\item Using this conditional association measure, we present a procedure to
		manually construct DAGs by ranking violation of implied CIs in the 
		model. This iterative procedure allows users to integrate their
		domain knowledge in the model construction procedure.
	\item Lastly, to allow researchers to easily apply this method on their
		datasets, we present a web-browser based interactive tool as well
		as implementation in a Python package.
\end{enumerate}

The rest of the paper is structured as follows. In
Section~\ref{sec:background}, we give a background on the commonly used
measures of association for various data types. In
Section~\ref{sec:mixed_association}, we present our generalized measure of
conditional association for mixed data. Section~\ref{sec:modification} provides
details on the procedure for using the measure of association for constructing
DAGs. Section~\ref{sec:web} presents our web-browser based tool and lastly in
Section~\ref{sec:empirical}, we show some empirical results to compare this
manual DAG construction approach to automated algorithms.

@@

\begin{figure}[t!]
	\begin{subfigure}{0.17 \textwidth}
		\includegraphics[page=1]{example.pdf}
		\caption{True DAG}
	\end{subfigure}%
	\begin{subfigure}{0.17 \textwidth}
		\includegraphics[page=2]{example.pdf}
		\caption{No edges}
	\end{subfigure}%
	\begin{subfigure}{0.17 \textwidth}
		\includegraphics[page=3]{example.pdf}
		\caption{Orient $ X_1 \rightarrow X_3 $}
	\end{subfigure}
	\begin{subfigure}{0.17 \textwidth}
		\includegraphics[page=4]{example.pdf}
		\caption{Orient $ X_2 \rightarrow X_3 $}
	\end{subfigure}%
	\begin{subfigure}{0.17 \textwidth}
		\includegraphics[page=5]{example.pdf}
		\caption{Orient $ X_3 \rightarrow X_4 $}
	\end{subfigure}%
	\begin{subfigure}{0.17 \textwidth}
		\includegraphics[page=6]{example.pdf}
		\caption{Orient $ X_3 \rightarrow X_5 $}
	\end{subfigure}
	\caption{An example showing a greedy approach for causal discovery using the mixed data measure of association. We start with an empty graph (b) where all the variables in the model are associated with each other. The strength of association is denoted using the thickness of the dotted edges. We select the pair of variables with the highest association, for example, $X_1$ and $X_3$, and orient it based on our domain knowledge. We continue the process till we have a DAG.
	}
\end{figure}

@@

\begin{figure}[t!]
	\centering
	\includegraphics[scale=0.4]{../code/plots/web_tool_full.png}
	\caption{A screenshot of the web tool for constructing the model. Users
		can upload their dataset after which the tool creates an empty
		graph and shows all pair of variables which are associated in
		the model using undirected red edges with the strength of
		association represented using edge width. Users can then add
		edges to the model (shown in green). Unnecessary edges are shown in 
		black.}
\end{figure}

@@

\begin{figure}[t!]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/shd_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/sid_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\caption{Comparison of PC and Hill-Climb Search algorithms against
		manually drawn DAGs using the assistance method. As PC and
		Hill-Climb Search return the Markov Equivalence Class (MEC), we
		use the best and worst scoring orientation of the MEC to get
		the range of values. The human drawn values are done for
		multiple accuracy values of the oracle.}
\end{figure}

\begin{figure}[t!]
	\begin{subfigure}{0.25\textwidth}
		\centering
		\includegraphics{../code/fig4/unexplained_effect.pdf}
		\caption{}
	\end{subfigure}%
	\begin{subfigure}{0.25\textwidth}
		\centering
		\includegraphics{../code/fig4/ll.pdf}
		\caption{}
	\end{subfigure}
	\caption{Plots showing improvements in fit of the model over iterations
	on the Adult income dataset. For the expert simulation an LLM is used
	for the edge orientations. (a) Shows using total unexplained effect that is the
	sum of effect size between variables that don't have an edge between them. (b)
	Shows the log-likelihood of the overall model.
	}
\end{figure}

@@

\begin{figure}[t!]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\begin{Verbatim}[fontsize=\footnotesize]
You are an expert in Social Science. Following
are the descriptions of two variables:
